---
title: "Gaussian Component-wise Boosting"
output: 
  rmarkdown::pdf_document:
    highlight: default
vignette: >
  %\VignetteIndexEntry{Gaussian Component-wise Boosting}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
header-includes:
  - \renewcommand{\familydefault}{\sfdefault}
  - \usepackage{xcolor}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
author:
  Luca Messerschmidt
  Bjarne Herbst
  Benedict B. Schnitzler
date: "2026-05-21"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">"
)
options(max.print = 100)
```

# Overview


This vignette introduces the basic workflow for fitting Gaussian component-wise
boosting models using **asp26boost**.

Boosting is a stage-wise optimization procedure that combines many simple model
updates into a flexible prediction model. From a statistical perspective,
boosting can be viewed as an iterative empirical risk minimization algorithm.
Starting from an initial estimate, the model is repeatedly improved by fitting
simple base learners to the current residual information and updating only the
best-performing model component at each iteration. Early stopping acts as a form
of regularization and often leads to sparse models by selecting only a subset of
available predictors.

For Gaussian response models, component-wise boosting minimizes the negative log-likelihood. Beginning with an intercept-only model, each predictor is fitted
separately to the current residuals. The predictor yielding the largest
reduction is selected and updated by a shrunken step
determined by the learning rate. Repeating this procedure yields coefficient
estimates, variable selection, and regularization simultaneously. In the
implementation used here, separate predictor sets can be specified for the mean
(`mu`) and scale (`sigma`) components via the arguments `X` and `Z`.

The implementation provided by **asp26boost** follows this component-wise
boosting framework and offers methods for model fitting, prediction,
summarization, and visualization.

# Installation

```{r include = TRUE}
devtools::load_all()
```


```{r include = FALSE, eval = FALSE}
devtools::install_gitlab(
  repo = "bjarne.herbst01/asp26boost",
  host = "https://gitlab.gwdg.de"
)
```


# Simulated Sample Data

To illustrate the basic functionality, we generate a synthetic regression
dataset with a sparse underlying signal.

```{r}
set.seed(27092001)

n <- 1000  # number of observations
p <- 10    # number of covariates

X <- matrix(
  rnorm(n * p),
  nrow = n,
  ncol = p
)

colnames(X) <- paste0("x", seq_len(p))

# Heteroskedastic scale model using the same three predictors
beta_mu <- c(2, -1.5, 1, rep(0, p - 3))
mu <- as.numeric(X %*% beta_mu)

beta_sigma <- c(0.4, -0.3, 0.2, rep(0, p - 3))
sigma <- exp(as.numeric(X %*% beta_sigma) / 2)

y <- rnorm(n, mean = mu, sd = sigma)


sample_data <- data.frame(
  y = as.numeric(y),
  X
)

head(sample_data)
```

The response depends only on the first three predictors. The remaining
covariates contain noise and are included to illustrate the variable-selection
behavior of component-wise boosting.

# Model Fitting

Gaussian component-wise boosting models are fitted using `boost_gaussian()`.

```{r}
fit <- boost_gaussian(
  X = sample_data[, -1],   # predictors for mu
  Z = sample_data[, -1],   # predictors for sigma
  y = sample_data$y,
  mstop = 100,
  nu_mu = 0.05,
  nu_sigma = 0.05
)
```

The most important tuning parameters are:

- `mstop`: number of boosting iterations.
- `nu_*`: learning rate controlling the size of each boosting update, defined seperately for the mean and scale.

Internally, predictors are standardized before fitting. The model is initialized
with the mean response and an initial scale estimate and updated iteratively by
selecting the predictor that provides the largest reduction in squared-error
loss.

Printing the fitted object provides a compact overview of the model.

```{r}
fit
```

Typical output includes:

- the original model call
- the tuning parameters specified during the model call
- The number of predictors, including which were (not) selected
- basic training-fit statistics
- the names of the selected variables and selection frequencies

# Model Summary

A more detailed summary can be obtained using `summary()`.

```{r}
summary(fit)
```

The summary object provides additional information on:

- summary of residuals
- coefficient estimates on the original predictor scale for selected predictors
- list of non-selected predictors (`coefficient = 0`)



# Prediction

Predictions can be computed using `predict()`.

```{r}
predict(fit)
```

When no additional data are supplied, fitted values for the training sample are
returned.

Predictions for new observations can be generated via the `newdata` argument.

```{r}
new_data <- sample_data[1:5, -1]

predict(
  fit,
  newdata_X = new_data,
  newdata_Z = new_data
)
```

The prediction method also supports evaluating models at earlier boosting
iterations.

```{r}
predict(
  fit,
  newdata_X = new_data,
  newdata_Z = new_data,
  mstop = 25
)
```

**Note**
For in-sample early-stopped predictions, the model must have been fitted with
`keep_path = TRUE`, which is the default when `keep_path` is unspecified.


# Model Diagnostics and Visualization

The package provides graphical diagnostics through the `plot()` method.

Coefficient trajectories across boosting iterations can be visualized using `plot()` method by specifying `type = "path"`.

```{r}
# For Location
plot(
  fit,
  type = "path",
  submodel = "mu",
  top_n = 8,
  scale = "original"
)
# And for scale
plot(
  fit,
  type = "path",
  submodel = "sigma",
  top_n = 8,
  scale = "original"
)
```

By default, the eight coefficients with the highest selection frequencies are displayed. Coefficients
may be shown either on the original (default) predictor scale or on the standardized
training scale.

Variable selection frequencies can be inspected by setting `type = "frequency"`.

```{r}
# For Location
plot(
  fit,
  type = "frequency",
  submodel = "mu",
  top_n = 8
)
# And for scale
plot(
  fit,
  type = "frequency",
  submodel = "sigma",
  top_n = 8
)
```

This visualization summarizes how often each predictor was selected during the
boosting process and provides a simple measure of variable importance. Again the number of Predictors displayed can be adjusted via `top_n`.

**Note**
To use the `plot()`diagnostics the model must be fitted with `keep_path = TRUE`, which is the default.



# Cross-Validation for Selecting the Stopping Iteration

A key tuning parameter in component-wise boosting is the number of boosting
iterations, denoted by `mstop`. While increasing the number of iterations
typically decreases the training error, excessively large values may lead to
overfitting. Consequently, early stopping serves as the primary regularization
mechanism in statistical boosting, encouraging sparsety in the selected variables.


A common strategy for selecting an appropriate stopping iteration is
cross-validation. The data is repeatedly split into training and validation
subsets. Models are fitted on the training folds and evaluated on the
corresponding validation folds. The value of `mstop` that minimizes the
estimated out-of-sample prediction error is then selected.

The function `cv_boost_gaussian()` implements k-fold cross-validation for
Gaussian component-wise boosting models.

```{r}
K = 5 # Number of folds during cross-validation

cv_fit <- cv_boost_gaussian(
  X = sample_data[, -1],
  Z = sample_data[, -1],
  y = sample_data$y,
  k = K,
  mstop = 1000,
  nu_mu = 0.05,
  nu_sigma = 0.05,
  seed = 27092001
)
```

# Inspecting Cross-Validation Results

The resulting object contains the selected variables in the optimal model,
the optimal `mstop`, and a final model refitted on the complete dataset using the optimal value of `mstop`.


```{r}
cv_fit
```

A more detailed summary can be obtained via:

```{r}
summary(cv_fit)
```

This gives additional information on the Residuals and the Coefficients.


# Visualizing the Cross-Validation Curve

The estimated validation error can be visualized using the plot method.

```{r}
plot(cv_fit)
```

The blue curve represents the mean validation error across folds as a function
of the boosting iteration. The vertical dashed line indicates the selected
stopping iteration minimizing the estimated prediction error.


# Prediction Using the Cross-Validated Model

Predictions from a cross-validated model can be generated directly via
`predict()`.

```{r}
predict(cv_fit)
```

When no new data is supplied, fitted values from the refitted model are
returned.

Predictions for new observations can be obtained by specifying `newdata`.

```{r}
new_data <- sample_data[1:5, -1]

predict(
  cv_fit,
  newdata_X = new_data,
  newdata_Z = new_data
)
```

Internally, predictions are delegated to the final boosting model fitted using
the optimal `mstop` value.

# Hyperparameter Grid Search

In some situations, both the stopping iteration `mstop` and the learning rate
`nu` may be treated as tuning parameters. The function `cv_boost_grid()`
evaluates combinations of candidate values using repeated cross-validation.

```{r eval=TRUE}
grid_fit <- cv_boost_grid(
  X = sample_data[, -1],
  Z = sample_data[, -1],
  y = sample_data$y,
  k = 5,
  mstop_grid = c(50, 100, 200),
  nu_mu_grid = c(0.01, 0.1, 0.3),
  nu_sigma_grid = c(0.01, 0.1, 0.3),
  seed = 27092001
)
```

```{r eval=TRUE}
grid_fit

```

The returned object contains the formula call and the best combination of hyperparameters.

# Concluding Remarks

This vignette introduced the basic workflow for Gaussian component-wise
boosting, including model fitting, prediction, visualization, and
cross-validation-based hyperparameter tuning.
