---
title: "Gaussian Component-wise Boosting"
output: 
  rmarkdown::pdf_document:
    highlight: default
vignette: >
  %\VignetteIndexEntry{Gaussian Component-wise Boosting}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
header-includes:
  - \renewcommand{\familydefault}{\sfdefault}
  - \usepackage{xcolor}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
author:
  Luca Messerschmidt
  Bjarne Herbst
  Benedict B. Schnitzler
date: "2026-05-21"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">"
)
options(max.print = 100)
```

# Overview

TODO: add citation

This vignette introduces the basic workflow for fitting Gaussian component-wise
boosting models using **asp26boost**.

Boosting is a stage-wise optimization procedure that combines many simple model
updates into a flexible prediction model. From a statistical perspective,
boosting can be viewed as an iterative empirical risk minimization algorithm.
Starting from an initial estimate, the model is repeatedly improved by fitting
simple base learners to the current residual information and updating only the
best-performing model component at each iteration. Early stopping acts as a form
of regularization and often leads to sparse models by selecting only a subset of
available predictors.

For Gaussian response models, component-wise boosting minimizes the squared-error
loss. Beginning with an intercept-only model, each predictor is fitted
separately to the current residuals. The predictor yielding the largest
reduction in residual sum of squares is selected and updated by a shrunken step
determined by the learning rate. Repeating this procedure yields coefficient
estimates, variable selection, and regularization simultaneously.

The implementation provided by **asp26boost** follows this component-wise
boosting framework and offers methods for model fitting, prediction,
summarization, and visualization.

# Installation

TBD
TODO, for now only to be functional
```{r include = FALSE}
devtools::load_all()
```




# Simulated Example Data

To illustrate the basic functionality, we generate a synthetic regression
dataset with a sparse underlying signal.

```{r}
set.seed(27092001)

n <- 1000  # number of observations
p <- 10    # number of covariates

X <- matrix(
  rnorm(n * p),
  nrow = n,
  ncol = p
)

colnames(X) <- paste0("x", seq_len(p))

beta <- c(2, -1.5, 1, rep(0, p - 3))

y <- X %*% beta + rnorm(n)

sample_data <- data.frame(
  y = as.numeric(y),
  X
)

head(sample_data)
```

The response depends only on the first three predictors. The remaining
covariates contain noise and are included to illustrate the variable-selection
behavior of component-wise boosting.

# Model Fitting

Gaussian component-wise boosting models are fitted using `boost_gaussian()`.

```{r}
fit <- boost_gaussian(
  x = sample_data[, -1],
  y = sample_data$y,
  mstop = 100,
  nu = 0.05
)
```

The most important tuning parameters are:

- `mstop`: number of boosting iterations.
- `nu`: learning rate controlling the size of each boosting update.

Internally, predictors are standardized before fitting. The model is initialized
with the mean response and updated iteratively by selecting the predictor that
provides the largest reduction in squared-error loss.

Printing the fitted object provides a compact overview of the model.

```{r}
fit
```

Typical output includes:

- the original model call
- the tuning parameters specified during the model call
- The number of predictors, including which were (not) selected
- basic training-fit statistics
- the names of the selected variables and selection frequencies

# Model Summary

A more detailed summary can be obtained using `summary()`.

```{r}
summary(fit)
```

The summary object contains information on:

- the original model call
- model dimensions
- tuning parameters
- summary of residuals
- basic training-fit statistics
- coefficient estimates on the original predictor scale for selected predictors
- list of non-selected predictors (`coefficient = 0`)



# Prediction

Predictions can be computed using `predict()`.

```{r}
predict(fit)
```

When no additional data are supplied, fitted values for the training sample are
returned.

Predictions for new observations can be generated via the `newdata` argument.

```{r}
new_data <- sample_data[1:5, -1]

predict(
  fit,
  newdata = new_data
)
```

The prediction method also supports evaluating models at earlier boosting
iterations.

```{r}
predict(
  fit,
  newdata = new_data,
  mstop = 25
)
```

**Note**
For in-sample early-stopped predictions, the model must have been fitted with
`keep_path = TRUE`, which is the default when `keep_path` is unspecified.


# Model Diagnostics and Visualization

The package provides graphical diagnostics through the `plot()` method.

Coefficient trajectories across boosting iterations can be visualized using the
path plot.

```{r}
plot(
  fit,
  type = "path"
)
```

By default, the most frequently selected variables are displayed. Coefficients
may be shown either on the original predictor scale or on the standardized
training scale.

Variable selection frequencies can be inspected using

```{r}
plot(
  fit,
  type = "frequency"
)
```

This visualization summarizes how often each predictor was selected during the
boosting process and provides a simple measure of variable importance.

**Note**
To use the `plot()`diagnostics the model must be fitted with `keep_path = TRUE`, which is the default.



# Cross-Validation for Selecting the Stopping Iteration

A key tuning parameter in component-wise boosting is the number of boosting
iterations, denoted by `mstop`. While increasing the number of iterations
typically decreases the training error, excessively large values may lead to
overfitting. Consequently, early stopping serves as the primary regularization
mechanism in statistical boosting.

TODO: Can be seen with x10 and x8?


A common strategy for selecting an appropriate stopping iteration is
cross-validation. The data are repeatedly split into training and validation
subsets. Models are fitted on the training folds and evaluated on the
corresponding validation folds. The value of `mstop` that minimizes the
estimated out-of-sample prediction error is then selected.

The function `cv_boost_gaussian()` implements k-fold cross-validation for
Gaussian component-wise boosting models.

```{r}
K = 5 # Number of folds during cross-validation

cv_fit <- cv_boost_gaussian(
  x = sample_data[, -1],
  y = sample_data$y,
  k = K,
  mstop = 1000,
  nu = 0.05,
  seed = 27092001
)
```

The resulting object contains the cross-validation errors for all boosting
iterations, the selected stopping iteration, and a final model refitted on the
complete dataset using the optimal value of `mstop`.

# Inspecting Cross-Validation Results

```{r}
cv_fit
```

A more detailed summary can be obtained via:

```{r}
summary(cv_fit)
```

The summary includes:

TODO

# Visualizing the Cross-Validation Curve

The estimated validation error can be visualized using the plot method.

```{r}
plot(cv_fit)
```

The blue curve represents the mean validation error across folds as a function
of the boosting iteration. The vertical dashed line indicates the selected
stopping iteration minimizing the estimated prediction error.

This diagnostic plot is useful for assessing whether model performance has
stabilized and whether the selected value of `mstop` lies in a region of
relatively low validation error.

# Prediction Using the Cross-Validated Model

Predictions from a cross-validated model can be generated directly via
`predict()`.

```{r}
predict(cv_fit)
```

When no new data are supplied, fitted values from the refitted model are
returned.

Predictions for new observations can be obtained using:

```{r}
new_data <- sample_data[1:5, -1]

predict(
  cv_fit,
  newdata = new_data
)
```

Internally, predictions are delegated to the final boosting model fitted using
the selected stopping iteration.

# Hyperparameter Grid Search

In some situations, both the stopping iteration `mstop` and the learning rate
`nu` may be treated as tuning parameters. The function `cv_boost_grid()`
evaluates combinations of candidate values using repeated cross-validation.

```{r eval=FALSE}
grid_fit <- cv_boost_grid(
  x = sample_data[, -1],
  y = sample_data$y,
  k = 5,
  mstop_grid = c(50, 100, 200),
  nu_grid = c(0.01, 0.1, 0.3),
  seed = 27092001
)
```
TODO `cv_boost_grid()` funktioniert nicht


The returned object contains:

- the best-performing parameter combination,
- cross-validation results for all evaluated combinations,
- a final model fitted using the selected hyperparameters.

Grid search can be useful when there is substantial uncertainty regarding the
appropriate learning rate or when predictive performance is the primary
objective.

# Concluding Remarks

This vignette introduced the basic workflow for Gaussian component-wise
boosting, including model fitting, prediction, visualization, and
cross-validation-based hyperparameter tuning.
